---
output:
  html_document:
    toc: true                               # table of contents
    toc_float: true                         # float the table of contents to the left of the main document content
    toc_depth: 3                            # header levels 1,2,3
    theme: default
    number_sections: false                  # add section numbering to headers
    df_print: paged                         # tables are printed as an html table with support for pagination over rows and columns
    highlight: pygments
    pdf_document: true
  pdf_document:
    toc: yes
date: "`r Sys.Date()`"
params:
  meta: NULL
  input_dir: NULL
  artifact_dir: NULL
  cpus: 1 
  workflow_version: v0.0.0
  workflow_commit: 0
  feature_type: gene
  observation_type: sample
  raw_matrix: null                                            # e.g. 0_salmon.merged.gene_counts.tsv
  normalised_matrix: null
  variance_stabilised_matrix: null                            # e.g. test_files/3_treatment-WT-P23H.vst.tsv
  contrasts_file: null                                        # e.g. GSE156533.contrasts.csv
  samples_file: null                                          # GSE156533.samplesheet.csv
  features_file: null
  differential_table: file.csv
  n_features: 500
  clustering_method: "ward.D2"
  cor_method: "spearman"
  versions_file: null                                         # e.g 17_software_versions.yml
  logo: null
  css: null 
---

<!-- Load libraries -->

```{r, include=FALSE}
library(knitr)
library(yaml)
library(shinyngs)
library(plotly)
```

```{r, include=FALSE}
save.image()
versions <- unlist(yaml.load_file(file.path(params$input_dir, params$versions_file)), recursive = FALSE)
```

---
title:  "<img src=\"`r file.path(params$input_dir, params$logo)`\" style=\"float: left;\"/>Differential `r params$feature_type` abundance report"
subtitle: differentialabundance workflow version `r versions[["Workflow.nf-core/differentialabundance"]]`
html_document:
    css_file: "`r file.path(params$input_dir, params$css)`"
---

<!-- set notebook defaults -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


<!-- Load input data -->


# Abstract

This report summarises differential `r params$feature_type` analysis as performed by the nf-core differential abundance pipeline.  

# Data

A summary of sample metadata is below:

```{r, echo=FALSE, results='asis'}
samples <- read_metadata(params$samples_file)
rownames(samples) <- samples[,1]
sample_name_col <- colnames(samples)[1]

contrasts <- read_metadata(params$contrasts_file)

# Pick the first variable used in contrasts for coloring purposes
main_grouping_variable <- contrasts$variable[1]
groupColorScale <- makeColorScale(length(unique(samples[[main_grouping_variable]])), palette = "Set1")

display_columns <- c(colnames(samples)[1], unique(contrasts$variable))
minimal_fetchngs_cols <- c('sample', 'sample_title', 'strandedness', 'library_strategy', 'scientific_name')

# If the data came via fetchngs then we can infer a couple of things about the most useful columns

if (all(minimal_fetchngs_cols %in% colnames(samples))){
  display_columns <- c(display_columns, minimal_fetchngs_cols)
  sample_name_col <- 'sample_title'
}else{
  not_wide_columns <- colnames(samples)[which(apply(samples, 2, function(x) max(nchar(x))) <= 20)]
  display_columns <- c(display_columns, minimal_fetchngs_cols[1:min(c(5,ncol(samples)))])
}
samples_to_print <- samples[,unique(display_columns)]
colnames(samples_to_print) <- prettifyVariablename(colnames(samples_to_print))
kable(samples_to_print)

```

Comparisons were made between sample groups defined using using this metadata, as described in the following table of contrasts:

```{r, echo=FALSE, results='asis'}
contrasts_to_print <- contrasts
colnames(contrasts_to_print) <- prettifyVariablename(colnames(contrasts_to_print))
kable(contrasts_to_print, caption = paste0("Table of contrasts. Positive log fold changes will come from ", params$feature_type, "s with higher values in the 'treatment' group."))
```

# Results

## Counts

## Preprocessing

## Exploratory

### Abundance value distributions

The following plots show the abundance value distributions of input matrices. A log(2) transform is applied where required.

```{r, include=FALSE}
assay_files <- list(
  raw = params$raw_matrix,
  normalised = params$normalised_matrix,
  variance_stabilised = params$variance_stabilised_matrix
)
assay_files <- Filter(Negate(is.null), assay_files)
assay_data <- lapply(assay_files, function(x) {
  mat <- read_matrix(
    x,
    sample_metadata = samples,
    row.names = 1
  )
  colnames(mat) <- samples[[sample_name_col]][match(colnames(mat), rownames(samples))]
  mat
})

# Now we can rename the samples rows using the title field
rownames(samples) <- samples[[sample_name_col]]
```
#### Box plots

```{r, echo=FALSE, results='asis'}

ggplot_boxplot(
  assay_data, 
  experiment = samples, 
  colorby = main_grouping_variable, 
  expressiontype = paste("count per", params$feature_type), 
  palette = groupColorScale
)
```

#### Density plots

```{r, echo=FALSE, results='asis'}
plotly_densityplot(
  assay_data, 
  experiment = samples, 
  colorby = main_grouping_variable, 
  expressiontype = paste("count per", params$feature_type), 
  palette = groupColorScale
)
```

### Sample relationships

#### Principal components plots

```{r, echo=FALSE, results='asis'}
# This is due to a bug in shinyngs, will be corrected
opt <<- list(ntop = params$n_features)
pca_data <- compilePCAData(assay_data$variance_stabilised)

for (contrast_variable in unique(contrasts$variable)){
  plotdata <- pca_data$coords
  plotdata$colorby <- factor(
    samples[[contrast_variable]], 
    levels = unique(samples[[contrast_variable]])
  )

  # Make plotting data combining PCA coords with coloring groups etc

  plotdata$name <- rownames(plotdata)
  percentVar <- pca_data$percentVar
  labels <- paste0(colnames(plotdata), " (", sprintf("%.1f", percentVar), "%)")
  ncats <- length(unique(plotdata$colorby))

  plot_types <- list("2" = "scatter", "3" = "scatter3d")

  for (d in names(plot_types)) {
  
    # Default plot args whatever we're doing

    plot_args <- list(
      x = pca_data$coords[, 1],
      y = pca_data$coords[, 2],
      xlab = labels[1],
      ylab = labels[2],
      colorby = plotdata$colorby,
      plot_type = plot_types[[d]],
      palette = groupColorScale,
      legend_title = prettifyVariablename(params$contrast_variable),
      labels = plotdata$name,
      show_labels = TRUE
    )
    
    if (d == "3") {
      plot_args$z <- pca_data$coords[, 3]
      plot_args$zlab <- labels[3]
    }

    print(htmltools::tagList(do.call("plotly_scatterplot", plot_args)))
  }
}
```

#### Clustering dendrograms

```{r, echo=FALSE, results='asis'}
variable_genes <- selectVariableGenes(matrix = assay_data$variance_stabilised, ntop = params$n_features)

p <- clusteringDendrogram(
  2^assay_data$variance_stabilised[variable_genes, ],
  samples[, main_grouping_variable, drop = FALSE],
  colorby = main_grouping_variable,
  cor_method = params$cor_method,
  plot_title = paste0(
    "Sample clustering dendrogram, ", 
    params$n_features, 
    " most variable ", 
    params$feature_type, 
    "s\n(", params$clustering_method, " clustering, ", params$cor_method, " correlation)"),
  cluster_method = params$clustering_method,
  palette = groupColorScale,
  labelspace = 0.25
)

# Defaults in shinyngs make the text in this plot a bit big for the report, so
# scale it down a bit
print(p, vp=grid::viewport(gp=grid::gpar(cex=0.7)))

```

## Differential

# Methods



## Exploratory 





## Differential

# Appendices

## Software versions

```{r, echo=FALSE, results='asis'}
versions_table <- data.frame(do.call(rbind, strsplit(names(versions), split = '\\.')), unlist(versions))
colnames(versions_table) <- c('Component', 'Software', 'Version')
kable(versions_table, row.names = FALSE)
```

# Citations

 <!-- TODO: Check citations-->

If you use nf-core/differentialabundance for your analysis, please cite it using the following DOI: TODO

Please also cite the `nf-core` and `Nextflow` publications:

- [nf-core](https://pubmed.ncbi.nlm.nih.gov/32055031/)

  > Ewels PA, Peltzer A, Fillinger S, Patel H, Alneberg J, Wilm A, Garcia MU, Di Tommaso P, Nahnsen S. The nf-core framework for community-curated bioinformatics pipelines. Nat Biotechnol. 2020 Mar;38(3):276-278. doi: 10.1038/s41587-020-0439-x. PubMed PMID: 32055031.
- [Nextflow](https://pubmed.ncbi.nlm.nih.gov/28398311/)

  > Di Tommaso P, Chatzou M, Floden EW, Barja PP, Palumbo E, Notredame C. Nextflow enables reproducible computational workflows. Nat Biotechnol. 2017 Apr 11;35(4):316-319. doi: 10.1038/nbt.3820. PubMed PMID: 28398311.
In addition, citations for the tools and data used in this pipeline are as follows:

- [Anaconda](https://anaconda.com)

  > Anaconda Software Distribution. Computer software. Vers. 2-2.4.0. Anaconda, Nov. 2016. Web.
- [Bioconda](https://pubmed.ncbi.nlm.nih.gov/29967506/)

  > Grüning B, Dale R, Sjödin A, Chapman BA, Rowe J, Tomkins-Tinch CH, Valieris R, Köster J; Bioconda Team. Bioconda: sustainable and comprehensive software distribution for the life sciences. Nat Methods. 2018 Jul;15(7):475-476. doi: 10.1038/s41592-018-0046-7. PubMed PMID: 29967506.
- [BioContainers](https://pubmed.ncbi.nlm.nih.gov/28379341/)

  > da Veiga Leprevost F, Grüning B, Aflitos SA, Röst HL, Uszkoreit J, Barsnes H, Vaudel M, Moreno P, Gatto L, Weber J, Bai M, Jimenez RC, Sachsenberg T, Pfeuffer J, Alvarez RV, Griss J, Nesvizhskii AI, Perez-Riverol Y. BioContainers: an open-source and community-driven framework for software standardization. Bioinformatics. 2017 Aug 15;33(16):2580-2582. doi: 10.1093/bioinformatics/btx192. PubMed PMID: 28379341; PubMed Central PMCID: PMC5870671.
- [Docker](https://dl.acm.org/doi/10.5555/2600239.2600241)

- [Singularity](https://pubmed.ncbi.nlm.nih.gov/28494014/)

  > Kurtzer GM, Sochat V, Bauer MW. Singularity: Scientific containers for mobility of compute. PLoS One. 2017 May 11;12(5):e0177459. doi: 10.1371/journal.pone.0177459. eCollection 2017. PubMed PMID: 28494014; PubMed Central PMCID: PMC5426675.
