---
params:
    args: ""

output:
  html_document:
    toc: true                               # table of contents
    toc_float: true                         # float the table of contents to the left of the main document content
    toc_depth: 3                            # header levels 1,2,3
    theme: default
    number_sections: false                  # add section numbering to headers
    df_print: paged                         # tables are printed as an html table with support for pagination over rows and columns
    css: ./nf-core_style.css
    highlight: pygments
    pdf_document: true
  html_notebook:
    toc: yes
---


---
title: "Differential abundance analysis"
subtitle: "Clonal abundance, diversity and V-family gene usage"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---


```{r setup, include=FALSE}
library(ggplot2)

theme_set(theme_bw(base_family = "ArialMT") +
            theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(family="ArialMT")))
knitr::opts_chunk$set(echo = FALSE)
```
***



```{r include = FALSE}
p_ca <- ggplot(mpg, aes(displ, hwy, colour = class)) +
  geom_point()

ggsave(plot=p_ca, filename = "bla.pdf", device="pdf", width = 25, height = 10, units="cm")
ggsave(plot=p_ca, filename = "bla.png", device="png", width = 25, height = 10, units="cm")
write("AAAAAAAAAAAA", file = "lalalala.txt")
```


# Citations

 <!-- TODO: Check citations-->

If you use nf-core/airrflow for your analysis, please cite it using the following DOI: [10.5281/zenodo.3607408](https://doi.org/10.5281/zenodo.3607408)

Please also cite the `nf-core` and `Nextflow` publications:

- [nf-core](https://pubmed.ncbi.nlm.nih.gov/32055031/)

  > Ewels PA, Peltzer A, Fillinger S, Patel H, Alneberg J, Wilm A, Garcia MU, Di Tommaso P, Nahnsen S. The nf-core framework for community-curated bioinformatics pipelines. Nat Biotechnol. 2020 Mar;38(3):276-278. doi: 10.1038/s41587-020-0439-x. PubMed PMID: 32055031.

- [Nextflow](https://pubmed.ncbi.nlm.nih.gov/28398311/)

  > Di Tommaso P, Chatzou M, Floden EW, Barja PP, Palumbo E, Notredame C. Nextflow enables reproducible computational workflows. Nat Biotechnol. 2017 Apr 11;35(4):316-319. doi: 10.1038/nbt.3820. PubMed PMID: 28398311.

In addition, citations for the tools and data used in this pipeline are as follows:

- [FastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/)

- [MultiQC](https://pubmed.ncbi.nlm.nih.gov/27312411/)

  > Ewels P, Magnusson M, Lundin S, Käller M. MultiQC: summarize analysis results for multiple tools and samples in a single report. Bioinformatics. 2016 Oct 1;32(19):3047-8. doi: 10.1093/bioinformatics/btw354. Epub 2016 Jun 16. PubMed PMID: 27312411; PubMed Central PMCID: PMC5039924.

## Software packaging/containerisation tools

- [Anaconda](https://anaconda.com)

  > Anaconda Software Distribution. Computer software. Vers. 2-2.4.0. Anaconda, Nov. 2016. Web.

- [Bioconda](https://pubmed.ncbi.nlm.nih.gov/29967506/)

  > Grüning B, Dale R, Sjödin A, Chapman BA, Rowe J, Tomkins-Tinch CH, Valieris R, Köster J; Bioconda Team. Bioconda: sustainable and comprehensive software distribution for the life sciences. Nat Methods. 2018 Jul;15(7):475-476. doi: 10.1038/s41592-018-0046-7. PubMed PMID: 29967506.

- [BioContainers](https://pubmed.ncbi.nlm.nih.gov/28379341/)

  > da Veiga Leprevost F, Grüning B, Aflitos SA, Röst HL, Uszkoreit J, Barsnes H, Vaudel M, Moreno P, Gatto L, Weber J, Bai M, Jimenez RC, Sachsenberg T, Pfeuffer J, Alvarez RV, Griss J, Nesvizhskii AI, Perez-Riverol Y. BioContainers: an open-source and community-driven framework for software standardization. Bioinformatics. 2017 Aug 15;33(16):2580-2582. doi: 10.1093/bioinformatics/btx192. PubMed PMID: 28379341; PubMed Central PMCID: PMC5870671.

- [Docker](https://dl.acm.org/doi/10.5555/2600239.2600241)

- [Singularity](https://pubmed.ncbi.nlm.nih.gov/28494014/)

  > Kurtzer GM, Sochat V, Bauer MW. Singularity: Scientific containers for mobility of compute. PLoS One. 2017 May 11;12(5):e0177459. doi: 10.1371/journal.pone.0177459. eCollection 2017. PubMed PMID: 28494014; PubMed Central PMCID: PMC5426675.

