---
output:
  html_document:
    toc: true                               # table of contents
    toc_float: true                         # float the table of contents to the left of the main document content
    toc_depth: 4                            # header levels 1,2,3
    theme: default
    number_sections: false                  # add section numbering to headers
    df_print: paged                         # tables are printed as an html table with support for pagination over rows and columns
    highlight: pygments
    pdf_document: true
  pdf_document:
    toc: yes
date: "`r Sys.Date()`"
params:
  meta: NULL
  input_dir: NULL
  artifact_dir: NULL
  cpus: 1 
  study_type: NULL
  study_name: NULL
  study_abundance_type: NULL
  report_file: NULL,
  report_title: NULL,
  report_author: NULL,
  report_description: NULL,
  observations_type: NULL
  observations: NULL                                          # GSE156533.samplesheet.csv
  observations_id_col: NULL
  observations_name_col: NULL
  features: NULL
  features_type: NULL
  features_id_col: NULL
  features_name_col: NULL
  features_metadata_cols: NULL 
  features_gtf_feature_type: NULL
  features_gtf_table_first_field: NULL
  features_log2_assays: NULL
  raw_matrix: null                                            # e.g. 0_salmon.merged.gene_counts.tsv
  normalised_matrix: null
  variance_stabilised_matrix: null                            # e.g. test_files/3_treatment-WT-P23H.vst.tsv
  contrasts_file: null                                        # e.g. GSE156533.contrasts.csv
  differential_table: file.csv
  affy_cel_files_archive: NULL
  affy_file_name_col: NULL
  affy_background: NULL
  affy_bgversion: NULL
  affy_destructive: NULL
  affy_cdfname: NULL
  affy_rm_mask: NULL
  affy_rm_outliers: NULL
  affy_rm_extra: NULL
  affy_build_annotation: NULL
  limma_ndups: NULL
  limma_spacing: NULL
  limma_block: NULL
  limma_correlation: NULL
  limma_method: NULL
  limma_proportion: NULL
  limma_stdev_coef_lim: NULL
  limma_trend: NULL
  limma_robust: NULL
  limma_winsor_tail_p: NULL
  limma_adjust_method: NULL
  limma_p_value: NULL
  limma_lfc: NULL
  limma_confint: NULL
  exploratory_n_features: null
  exploratory_clustering_method: null
  exploratory_cor_method: null
  exploratory_whisker_distance: null
  exploratory_mad_threshold: null
  exploratory_main_variable: null
  exploratory_assay_names: NULL
  exploratory_final_assay: NULL
  exploratory_palette_name: NULL
  versions_file: null                                         # e.g 17_software_versions.yml
  logo: null
  css: null
  citations: null
  filtering_min_samples: 1
  filtering_min_abundance: 1
  filtering_min_proportion: NULL
  filtering_grouping_var: NULL
  differential_file_suffix: NULL
  differential_feature_id_column: NULL
  differential_feature_name_column: NULL
  differential_fc_column: NULL
  differential_pval_column: NULL
  differential_qval_column: NULL
  differential_min_fold_change: NULL
  differential_foldchanges_logged: NULL
  differential_max_pval: NULL
  differential_max_qval: NULL
  differential_palette_name: NULL
  differential_subset_to_contrast_samples: NULL
  deseq2_test: NULL
  deseq2_fit_type: NULL
  deseq2_sf_type: NULL
  deseq2_min_replicates_for_replace: NULL
  deseq2_use_t: NULL
  deseq2_lfc_threshold: NULL
  deseq2_alt_hypothesis: NULL
  deseq2_independent_filtering: NULL
  deseq2_p_adjust_method: NULL
  deseq2_alpha: NULL
  deseq2_minmu: NULL
  deseq2_vs_method: NULL
  deseq2_shrink_lfc: NULL
  deseq2_cores: NULL
  deseq2_vs_blind: NULL
  deseq2_vst_nsub: NULL
  gsea_run: false
  gsea_nperm: NULL
  gsea_permute: NULL
  gsea_scoring_scheme: NULL    
  gsea_metric: NULL 
  gsea_sort: NULL
  gsea_order: NULL
  gsea_set_max: NULL
  gsea_set_min: NULL
  gsea_norm: NULL
  gsea_rnd_type: NULL
  gsea_make_sets: NULL
  gsea_median: NULL
  gsea_num: NULL
  gsea_plot_top_x: NULL
  gsea_rnd_seed: NULL
  gsea_save_rnd_lists: NULL
  gsea_zip_report: NULL
  gsea_chip_file: NULL 
  gsea_gene_sets: NULL
---

<!-- Load libraries -->

```{r, include=FALSE}
library(knitr)
library(yaml)
library(shinyngs)
library(plotly)
library(DT)
```

```{r include = FALSE}
# Load the datatables js
datatable(NULL)
```

```{r, include=FALSE}
versions <- unlist(yaml.load_file(file.path(params$input_dir, params$versions_file)), recursive = FALSE)
params_table <- data.frame(Parameter = names(unlist(params)), Value = unlist(params), row.names = NULL)

# We'll subset the params table for different report sections
make_params_table <- function(name, pattern = NULL, remove_pattern = FALSE){
  subparams <- params_table
  if (! is.null(pattern)){
    subparams <- subparams[grep(pattern, subparams$Parameter),]
  }
  if (remove_pattern){
    subparams$Parameter <- sub(pattern, '', subparams$Parameter)
  }
  
  if (nrow(subparams) > 10){
    dom <- 'tp'
  }else{
    dom <- 't'
  }
  
  print( htmltools::tagList(datatable(subparams, caption = paste("Parameters used for", name), rownames = FALSE, options = list(dom = dom)) ))
}

report_title <- paste0('Differential ',  params$features_type, ' abundance report', ifelse(is.null(params$report_title), '', paste0(': ', params$report_title)))
report_subtitle <- paste0(ifelse(is.null(params$report_author), '', paste0('By ', params$report_author, ', ')), 'differentialabundance workflow version', versions[["Workflow.nf-core/differentialabundance"]])
```

---
title:  "<img src=\"`r file.path(params$input_dir, params$logo)`\" style=\"float: left;\"/>`r report_title`"
subtitle: `r report_subtitle`
---

<!-- set notebook defaults -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- Include the CSS and set the logo -->

```{r, echo=FALSE}
htmltools::includeCSS(params$css)
```

```{r results="asis", echo=FALSE}
cat(paste0("
<style>
#TOC {
   background-image: url(\"", knitr::image_uri(params$logo), "\");
}
</style>
"))
```

<!-- Load input data -->

```{r, echo=FALSE}
observations <- read_metadata(file.path(params$input_dir, params$observations), id_col = params$observations_id_col)
if (! params$observations_name_col %in% colnames(observations)){
    stop(paste('Invalid observation name column specified: ', params$observations_name_col, paste0('(Valid values are: ', paste(colnames(observations), collapse=', '),')')))
}

if (! is.null(params$features)){
  features <- read_metadata(file.path(params$input_dir, params$features))
  if (! is.null(params$features_metadata_cols)){
    features <- features[,colnames(features) %in% simpleSplit(params$features_metadata_cols), drop = FALSE]
  }
}

contrasts <- read_metadata(file.path(params$input_dir, params$contrasts_file))
contrasts$blocking <- na.replace(contrasts$blocking, '')
if (! 'id' %in% colnames(contrasts)){
  contrasts$id <- apply(contrasts, 1, paste, collapse='_')
}

# Identify informative variables- those with a number of values greater than 1
# but less than N, with N being the number of observations. Make sure contrast
# variables are first in the list

informative_variables <- unique(c(contrasts$variable, chooseGroupingVariables(observations)))

# Remove any informative variables that group observations the same way
informative_variables <- informative_variables[ ! duplicated(lapply(structure(informative_variables, names= informative_variables), function(x) as.numeric(factor(observations[[x]], levels=unique(observations[[x]])))))]

assay_names <- simpleSplit(params$exploratory_assay_names)
names(assay_names) = assay_names
assay_files <- lapply(assay_names, function(x) params[[paste0(x, '_matrix')]])

# Set up vector of unlogged assay files (if any)
unlogged <- c()
features_log2_assays <- params$features_log2_assays

# For maxquant input override param as all assays are already logged by proteus
if (params$study_type == 'maxquant') {
  features_log2_assays <- ""
}

if (is.null(features_log2_assays)) {
  # Guess unlogged assays
  for (assay in c(1:length(assay_files))) {
    if (max(assay_files[[assay]]) > 20) {
      unlogged <- append(unlogged, assay)
    }
  }

} else {
  features_log2_assays <- gsub('\\]$', '', gsub('^\\[', '', features_log2_assays)) # Remove brackets from assay list

  if (features_log2_assays != "") {

    if (is_valid_positive_integer_vector(features_log2_assays)) {
      # Convert to list of assay positions
      unlogged <- unique(as.integer(simpleSplit(features_log2_assays)))
      invalid_assays <- unlogged[! unlogged %in% 1:length(assay_files)]

      if (length(invalid_assays) > 0){
        stop(paste0("Invalid assay numbers: ", paste(invalid_assays, collapse=', ')))
      }

    } else {
      # Last option is string of assay names, so get positions of those names in the assay list
      unlogged <- unique(simpleSplit(features_log2_assays))

      # Check if all names are valid
      invalid_assays <- unlogged[!(unlogged %in% names(assay_files))]
      if (length(invalid_assays) > 0) {
        stop(paste0(invalid_assays, " is/are not valid assay name(s). Valid assay names are: ", paste(names(assay_files), collapse=", "), ". Please check param --features_log2_assays."))
      }
      unlogged <- match(unlogged, names(assay_files))
    }

  }
}

assay_data <- lapply(c(1:length(assay_files)), function(x) {
  mat <- na.omit(
    read_matrix(
      assay_files[[x]],
      sample_metadata = observations,
      row.names = 1
    )
  )
  colnames(mat) <- observations[[params$observations_name_col]][match(colnames(mat), rownames(observations))]

  # If assay is in unlogged list, apply log2
  if (x %in% unlogged) {
    mat <- log2(mat+1)
  }
  mat
  
})

# Lapply does not go over the assay_files themselves anymore, so we need to specifically assign their names to the data
names(assay_data) <- names(assay_files)

# Now we can rename the observations rows using the title field
rownames(observations) <- observations[[params$observations_name_col]]

# Run PCA early so we can understand how important each variable is

pca_datas <- lapply(names(assay_data), function(assay_type){
  compilePCAData(assay_data[[assay_type]])
})
names(pca_datas) <- names(assay_data)

pca_vs_meta <- anova_pca_metadata(pca_datas[[params$exploratory_final_assay]]$coords, observations[,informative_variables, drop = FALSE], pca_datas[[params$exploratory_final_assay]]$percentVar)

# Show the variable with the tightest PC associations first
informative_variables <- rownames(pca_vs_meta)[order(pca_vs_meta[,1])]

# Pick the variable used for coloring purposes etc
if (params$exploratory_main_variable == 'contrasts'){
  main_grouping_variable <- contrasts$variable[1]
}else if (params$exploratory_main_variable == 'auto_pca'){
  main_grouping_variable <- informative_variables[1]
}else{
  if (! params$exploratory_main_variable %in% colnames(observations)){
    stop(paste('Invalid main variable specified: ', params$exploratory_main_variable))
  }
  main_grouping_variable <- params$exploratory_main_variable
}

# Make sure the main variable is shown first, with remaining shown in order of
# informativeness

informative_variables <- unique(c(main_grouping_variable, informative_variables))

groupColorScale <- makeColorScale(length(unique(observations[[main_grouping_variable]])), palette = params$exploratory_palette_name)
```

<!-- Read the differential results.
NOTE: differential results files are expected to have the pattern:

<variable>-<reference>-<target>-<blocking><differential_file_suffix>, e.g.
treatment-mCherry-hND6-batcheffect.deseq2.results.tsv

... where variable, reference, target and blocking come from the contrasts file
(with blocking being optional) and the suffix is defined in parameters.
-->

```{r, echo=FALSE}

differential_files <- lapply(contrasts$id, function(d){
 file.path(params$input_dir, paste0(gsub(' |;', '_', d), params$differential_file_suffix)) 
})

differential_results <- lapply(differential_files, function(diff_file){
  if (! file.exists(diff_file)){
    stop(paste("Differential file", diff_file, "does not exist"))
  }
  diff <- read_differential(
    diff_file, 
    feature_id_column = params$differential_feature_id_column,
    fc_column = params$differential_fc_column,
    pval_column = params$differential_pval_column,
    qval_column = params$differential_qval_column
  )
  
  # If fold changes are not logged already, log them (we assume they're logged
  # later on)
  
  if (! params$differential_foldchanges_logged){
    diff[[params$differential_fc_column]] <- log2(diff[[params$differential_fc_column]])
  }
  
  # Annotate differential tables if possible
  if (! is.null(params$features)){
    diff <- merge(features, diff, by.x = params$features_id_col, by.y = params$differential_feature_id_column)
  }
  diff
})
names(differential_results) <- contrasts$id
```

<!-- Calculate some summary statistics -->

```{r, echo=FALSE}

# Function to make friendly contrast name from contrast components, including optional bits

name_contrast <- function(i){
  contrast_name <- paste(contrasts$target[i], 'versus', contrasts$reference[i], 'in', contrasts$variable[i])
  contrast_vals <- contrasts[i,]
  populated <- colnames(contrasts)[! (is.na(contrast_vals) | contrast_vals == '' | is.null(contrast_vals))]
  optional <- setdiff(populated, c('id', 'target', 'reference', 'variable'))

  if (length(optional) > 0){
    optional_part <-  paste0('(', paste(paste(optional, contrasts[i,optional], sep=': '), collapse=', '), ')')
  }else{
    optional_part <- ''
  }

  paste(contrast_name, optional_part)
}

contrast_descriptions <- unlist(lapply(1:nrow(contrasts), function(x) name_contrast(x)))

# Check both adjusted and unadjusted p values

p_value_types <- list(Adjusted = params$differential_qval_column, Unadjusted = params$differential_pval_column)
p_value_thresholds <- list(Adjusted = params$differential_max_qval, Unadjusted = params$differential_max_pval)

sig_differential <- 
  lapply(names(p_value_types), function(pvt){
    diff <- lapply(
      1:nrow(contrasts),
      function(x){
        signif <- differential_results[[x]][,p_value_types[[pvt]] ] < p_value_thresholds[[pvt]]
        list(
          up = differential_results[[x]][which(
            differential_results[[x]][,params$differential_fc_column ] > log2(params$differential_min_fold_change) &
            signif
          ),],
          down = differential_results[[x]][which(
            differential_results[[x]][,params$differential_fc_column ] < log2(1/params$differential_min_fold_change) &
            signif 
          ),]
        ) 
      }
    )
    names(diff) <- contrast_descriptions
    diff
  })
names(sig_differential) <- names(p_value_types)

# Count the differential genes
differential_tables <- lapply(names(sig_differential), function(sd) do.call(rbind, lapply(sig_differential[[sd]], function(x) lapply(x, function(y) nrow(y)))))
names(differential_tables) <- names(sig_differential)
```

<!-- Write the report -->

# Abstract

This report summarises differential `r params$features_type` analysis as performed by the nf-core/differentialabundance pipeline. 

# Data

```{r, echo=FALSE, results='asis'}
cat(paste0("\n## ", ucfirst(params$observations_type), "s\n"))
```


A summary of `r params$observations_type` metadata is below:

```{r, echo=FALSE, results='asis'}
display_columns <- union(c(params$observations_id_col, unique(contrasts$variable)), informative_variables)
minimal_fetchngs_cols <- c('sample', 'sample_title', 'strandedness', 'library_strategy', 'scientific_name')

# If the data came via fetchngs then we can infer a couple of things about the most useful columns

if (all(minimal_fetchngs_cols %in% colnames(observations))){
  additional_useful_cols <- minimal_fetchngs_cols
}else{
  additional_useful_cols <- colnames(observations)[which(apply(observations, 2, function(x) max(nchar(x))) <= 20)] 
}

display_columns <- head(union(display_columns, additional_useful_cols), 5)

# Also add informative columns
display_columns <- unique(c(display_columns, informative_variables))
observations_to_print <- observations[,unique(display_columns)]
colnames(observations_to_print) <- prettifyVariablename(colnames(observations_to_print))
print( htmltools::tagList(datatable(observations_to_print, caption = paste(ucfirst(params$observations_type), 'metadata'), rownames = FALSE, options = list(dom = 't')) ))

```

## Contrasts

Comparisons were made between `r params$observations_type` groups defined using using `r params$observation_type` metadata columns, as described in the following table of contrasts:

```{r, echo=FALSE, results='asis'}
contrasts_to_print <- contrasts
colnames(contrasts_to_print) <- prettifyVariablename(colnames(contrasts_to_print))
print( htmltools::tagList(datatable(contrasts_to_print, caption = paste0("Table of contrasts"), rownames = FALSE, options = list(dom = 't')) ))
```

# Results

## Counts

Input was a matrix of `r nrow(assay_data$raw)` `r params$features_type`s for `r ncol(assay_data$raw)` `r params$observations_type`s`r ifelse(nrow(assay_data$normalised) < nrow(assay_data$raw), paste0(', reduced to ', nrow(assay_data$normalised), ' ', params$features_type, 's after filtering for low abundance'), '')`.

## Exploratory analysis

### Abundance value distributions

The following plots show the abundance value distributions of input matrices. A log2 transformation is applied where not already performed.

```{r, include=FALSE}

```
#### Box plots

```{r, echo=FALSE, results='asis', fig.height=8}

p <- ggplot_boxplot(
  assay_data, 
  experiment = observations, 
  colorby = main_grouping_variable, 
  expressiontype = paste("count per", params$features_type), 
  palette = groupColorScale,
  whisker_distance = params$exploratory_whisker_distance,
  base_size=8
)
print(p)
```

Whiskers in the above boxplots show `r params$exploratory_whisker_distance` times the inter-quartile range.

#### Density plots

```{r, echo=FALSE, results='asis', fig.height=8}
plotly_densityplot(
  assay_data, 
  experiment = observations, 
  colorby = params$observations_name_col, 
  expressiontype = paste("count per", params$features_type),
  makeColorScale(length(unique(observations[[params$observations_id_col]])), palette = "Set1")
)
```

```{r, echo=FALSE, results='asis'}
cat(paste0("\n### ", ucfirst(params$observations_type), " relationships\n"))
```

#### Principal components plots {.tabset}

Principal components analysis was conducted based on the `r params$exploratory_n_features` most variable `r params$features_type`s. Each component was annotated with its percent contribution to variance. 

```{r, echo=FALSE, results='asis'}
for (assay_type in rev(names(assay_data))){
  
  pca_data <- pca_datas[[assay_type]]

  for (iv in informative_variables){
  
    cat(paste0("\n##### ", prettifyVariablename(assay_type), " (", iv, ")\n"))
  
    plotdata <- pca_data$coords
    plotdata$colorby <- factor(
      observations[[iv]], 
      levels = unique(observations[[iv]])
    )
    pcaColorScale <- makeColorScale(length(unique(observations[[iv]])), palette = params$exploratory_palette_name)

    # Make plotting data combining PCA coords with coloring groups etc

    plotdata$name <- rownames(plotdata)
    percentVar <- pca_data$percentVar
    labels <- paste0(colnames(plotdata), " (", sprintf("%.1f", percentVar), "%)")
    ncats <- length(unique(plotdata$colorby))

    plot_types <- list("2" = "scatter", "3" = "scatter3d")

    for (d in names(plot_types)) {
  
      # Default plot args whatever we're doing

      plot_args <- list(
        x = pca_data$coords[, 1],
        y = pca_data$coords[, 2],
        xlab = labels[1],
        ylab = labels[2],
        colorby = plotdata$colorby,
        plot_type = plot_types[[d]],
        palette = pcaColorScale,
        legend_title = prettifyVariablename(iv),
        labels = plotdata$name,
        show_labels = TRUE
      )
    
      if (d == "3") {
        plot_args$z <- pca_data$coords[, 3]
        plot_args$zlab <- labels[3]
      }

      print(htmltools::tagList(do.call("plotly_scatterplot", plot_args)))
    }
  }
}
```

#### Principal components/ metadata associations

For the variance stabilised matrix, an ANOVA test was used to determine assocations between continuous principal components and categorical covariates (including the variable of interest).

The resulting p values are illustrated below.

```{r, echo=FALSE, results='asis'}

# This is a little hack to work around a bug in d3heatmap with single-row data
# frames.
if (nrow(pca_vs_meta) == 1){
  plot_pca_meta <- rbind(pca_vs_meta, pca_vs_meta)
}else{
  plot_pca_meta <- pca_vs_meta
}

d3heatmap::d3heatmap(
  -log10(plot_pca_meta),
  Rowv = FALSE,
  dendrogram = 'none',
  cellnote = plot_pca_meta,
  cexCol = 0.8, 
  cexRow = 0.8,
  height = (100 + (15 * nrow(plot_pca_meta))),
  colors = colorRampPalette(
    rev(
      RColorBrewer::brewer.pal(n = 7, name = "RdYlBu")
    )
  )(100)
)

for (variable in rownames(pca_vs_meta)){
  sig_comps <- pca_vs_meta[variable,] < 0.1

  if (any(sig_comps)){
    min_sig_comp <- min(which(sig_comps))
  
    min_sig_comp_p <- sprintf("%.2f", pca_vs_meta[variable, min_sig_comp])
    cat(paste0('The variable \'', variable, '\' shows an association with ', colnames(pca_vs_meta)[min_sig_comp], ' (p = ', min_sig_comp_p,'). '))
  }
}
```

#### Clustering dendrograms {.tabset}

A hierarchical clustering of `r params$features_type`s was undertaken based on the top  `r params$exploratory_n_features` most variable `r params$features_type`s. Distances between `r params$features_type`s were estimated based on `r params$exploratory_cor_method` correlation, which were then used to produce a clustering via the `r params$exploratory_clustering_method` method with `hclust()` in R. 

```{r, echo=FALSE, results='asis'}
for (assay_type in rev(names(assay_data))){
  for (iv in informative_variables){
    cat(paste0("\n##### ", prettifyVariablename(assay_type), " (", iv, ")\n"))
    variable_genes <- selectVariableGenes(matrix = assay_data[[assay_type]], ntop = params$exploratory_n_features)

    dendroColorScale <- makeColorScale(length(unique(observations[[iv]])), palette = params$exploratory_palette_name)

    p <- clusteringDendrogram(
      2^assay_data[[assay_type]][variable_genes, ],
      observations[, iv, drop = FALSE],
      colorby = iv,
      cor_method = params$exploratory_cor_method,
      plot_title = paste0(
        paste0(params$observations_type," clustering dendrogram, "), 
        params$exploratory_n_features, 
        " most variable ", 
        params$features_type, 
        "s\n(", params$exploratory_clustering_method, " clustering, ", params$exploratory_cor_method, " correlation)"),
      cluster_method = params$exploratory_clustering_method,
      palette = dendroColorScale,
      labelspace = 0.25
    )
    # Defaults in shinyngs make the text in this plot a bit big for the report, so
    # scale it down a bit
    print(p, vp=grid::viewport(gp=grid::gpar(cex=0.7)))
    cat("\n")
  }
}
```

### Outlier detection {.tabset}

Outlier detection based on [median absolute deviation](https://wiki.arrayserver.com/wiki/index.php?title=CorrelationQC.pdf) was undertaken, the outlier scoring is plotted below.

```{r, echo=FALSE, results='asis', warning=FALSE}

# We can't look for ouliers in sets of less than 3 samples, so exclude variables
# unless the minimum group size is larger than that
iv_min_group_sizes <- unlist(lapply(informative_variables, function(x) min(table(observations[[x]]))))

foo <- lapply(informative_variables[iv_min_group_sizes > 2], function(iv){
  
  cat(paste("\n####", iv, "\n"))
  
  plotdata <-
  madScore(
    matrix = assay_data[[params$exploratory_final_assay]],
    sample_sheet = observations,
    groupby = iv
  )

  if (! is.null(plotdata)){
    mad_plot_args <- list(
        x = plotdata$group,
        y = plotdata$mad,
        color = plotdata$outlier,
        hline_thresholds = c("Outlier threshold" = params$exploratory_mad_threshold),
        palette = makeColorScale(2, palette = params$differential_palette_name),
        legend_title = "Outlier status",
        labels = rownames(plotdata),
        show_labels = TRUE,
        xlab = "Sample group",
        ylab = "MAD score"
    )

    print(htmltools::tagList(do.call("plotly_scatterplot", mad_plot_args)))

    outliers <- rownames(plotdata)[plotdata$outlier]

    if (length(outliers) == 0){
        cat(paste0("No outlying samples were detected in groups defined by ", iv,".\n"))
    }else{
        cat(paste0(length(outliers), ' possible outliers were detected in groups defined by ', iv ,': ', paste(outliers, collapse=', '), "\n"))
    }
  }
})

```

## Differential analysis

### Differential `r params$features_type` `r params$study_abundance_type` {.tabset}

```{r, echo=FALSE, results='asis'}
foo <- lapply(names(p_value_types), function(pvt){
  cat("\n#### ", pvt, "\n")
  print( htmltools::tagList(datatable(differential_tables[[pvt]], caption = paste0('Differential ',  params$features_type, " ", params$abundance_type, ' (target relative to reference)'), options = list(dom = 't'), rownames = TRUE) ))
  cat("\n")
})
```

```{r, echo=FALSE, results='asis', eval = FALSE}

differential_summary_string <- paste(
  paste(
  lapply(
    1:nrow(contrasts),
    function(x){
      paste0(
        "Contrast ", x, ' (', contrast_descriptions[x], ') ', "had ", differential_table[x,'up'], ' ', paste0(params$features_type, 's'), ' expressed significantly more highly in ', contrasts[x, 'target',], ' than ', contrasts[x, 'reference',], ' and ', differential_table[x,'down'], ' expressed at sifnificantly lower levels.'
      )
    }
  ),
  collapse = ' '
  )
)
cat(differential_summary_string)
```

### Differential `r params$features_type` details

```{r, echo=FALSE, results='asis'}
for (i in 1:nrow(contrasts)){
  cat("\n#### ", contrast_descriptions[i], "  {.tabset}\n")
  
  ## Make a volcano plot for the contrast first
  
  # Label features with symbol as well as identifier
  if (! is.null(params$features) && (! is.null(params$differential_feature_name_column)) ){
    label_col <- params$differential_feature_name_column
  }else{
    label_col <- params$differential_feature_id_column
  }

  # Get the full set of differential stats for this contrast, removing rows with
  # NAs in the fields we need.
  full_de <- differential_results[[i]]
  full_de <- subset(full_de, (! is.na(full_de[[params$differential_fc_column]])) & (! is.na(full_de[[params$differential_qval_colum]])) )
  
  # We'll color by whether features are differential according to supplied thresholds

  p_value_types <- list(Adjusted = params$differential_qval_column, Unadjusted = params$differential_pval_column)
  p_value_thresholds <- list(Adjusted = params$differential_max_qval, Unadjusted = params$differential_max_pval)
  
  for (pvt in names(p_value_types)){
    cat("\n##### ", pvt, " p values\n")
    pval_column <- p_value_types[[pvt]]
  
    full_de$differential_status <- FALSE
    full_de$differential_status[abs(full_de[[params$differential_fc_column]]) > log2(params$differential_min_fold_change) & full_de[[pval_column]] < p_value_thresholds[[pvt]]] <- TRUE

    # Define the thresholds we'll draw

    hline_thresholds = vline_thresholds = list()
    hline_thresholds[[paste(pval_column, '=', p_value_thresholds[[pvt]])]] = -log10(p_value_thresholds[[pvt]])
    vline_thresholds[[paste(params$differential_fc_column, '<-', log2(params$differential_min_fold_change))]] = -log2(params$differential_min_fold_change)
    vline_thresholds[[paste(params$differential_fc_column, '>', log2(params$differential_min_fold_change))]] = log2(params$differential_min_fold_change)

    plot_args <- list(
      x = full_de[[params$differential_fc_column]],
      y = -log10(full_de[[pval_column]]),
      colorby = full_de$differential_status,
      ylab = paste("-log(10)", pval_column),
      xlab = xlabel <- paste("higher in", contrasts$reference[i], "          <<", params$differential_fc_column, ">>           higher in", contrasts$target[i]) ,
      labels = full_de[[label_col]],
      hline_thresholds = hline_thresholds,
      vline_thresholds = vline_thresholds,
      show_labels = FALSE,
      legend_title = "Differential status",
      palette = makeColorScale(2, params$differential_palette_name)
    )
  
    # Let's equalize the axes
    max_fc <- max(abs(full_de[[params$differential_fc_column]])) * 1.1
    p <- do.call(plotly_scatterplot, plot_args) %>%
      layout(xaxis = list(range=list(-max_fc, max_fc)))
 
    print(htmltools::tagList(p))
  
    ## ... then show tables of the up/ down genes
      
    for (dir in c('up', 'down')){
      contrast_de <- sig_differential[[pvt]][[i]][[dir]]
      cols_to_round <- c(params$differential_fc_column, params$differential_pval_column, params$differential_qval_column) 
      contrast_de[, cols_to_round] <- signif(contrast_de[, cols_to_round], 8)

      colnames(contrast_de) <- prettifyVariablename(colnames(contrast_de))
        
      if (nrow(contrast_de) > 0){
        print( htmltools::tagList(datatable(contrast_de, caption = paste('Differential genes', dir, 'in', contrast_descriptions[i], " (check", differential_files[[i]], "for more detail)"), rownames = FALSE) ))
      }else{
        cat(paste0("No significantly differential '", dir, "' genes.\n\n"))
      }
    }
  }
  
}
```

<!-- Gene set analysis results -->

```{r, echo=FALSE, results='asis'}
possible_gene_set_methods <- c('gsea')
if (any(unlist(params[paste0(possible_gene_set_methods, '_run')]))){
  cat("\n### Gene set analysis\n")

  for (gene_set_method in possible_gene_set_methods){
    if (unlist(params[paste0(gene_set_method, '_run')])){
      cat("\n### ", toupper(gene_set_method) ," {.tabset}\n")
      
      for (gmt_file in simpleSplit(params$gsea_gene_sets)) {
        gmt_name <- basename(tools::file_path_sans_ext(gmt_file))

        cat("\n#### ", gmt_name ," {.tabset}\n")
        reference_gsea_tables <- paste0(contrasts$id, ".", gmt_name, '.gsea_report_for_', contrasts$reference, '.tsv')
        target_gsea_tables <- paste0(contrasts$id, ".", gmt_name, '.gsea_report_for_', contrasts$target, '.tsv')

        for (i in 1:nrow(contrasts)){
          cat("\n##### ", contrast_descriptions[i], "\n")

          target_gsea_results <- read_metadata(target_gsea_tables[i])[,c(-2,-3)]
          print( htmltools::tagList(datatable(target_gsea_results, caption = paste0("\nTarget (", contrasts$target[i], ")\n"), rownames = FALSE) ))

          ref_gsea_results <- read_metadata(reference_gsea_tables[i])[,c(-2,-3)]
          print( htmltools::tagList(datatable(ref_gsea_results, caption = paste0("\nReference (", contrasts$reference[i], ")\n"), rownames = FALSE) ))
        }
      }
    }
  }
}
```

# Methods

## Filtering

```{r, echo=FALSE, results='asis'}
make_params_table('feature-wise filtering', 'filtering_', remove_pattern = TRUE)
```

```{r, echo=FALSE, results='asis'}
filtering_string <- paste0('Filtering was carried out by selecting ', params$features_type, 's with an abundance of at least ', params$filtering_min_abundance)

if (is.null(params$filtering_grouping_var)){
  if (is.null(params$filtering_min_proportion)){
    filtering_string <- paste0(filtering_string, ' in at least ', params$filtering_min_samples, ' ', params$observations_type, 's.')
  }else{
    filtering_string <- paste0(filtering_string, ' in at least a proportion of ', params$filtering_min_proportion, ' of ', params$observations_type,'s.')
  }
}else{
  if (is.null(params$filtering_min_proportion)){
    filtering_string <- paste0(filtering_string, ' in at least the number of ', params$observations_type, 's corresponding to the smallest group size defined by the grouping variable "', params$filtering_grouping_var, '".')
  }else{
    filtering_string <- paste0(filtering_string, ' in at least a proportion of ', params$filtering_min_proportion, ' of the number of ', params$observations_type,'s corresponding to the smallest group size defined by the grouping variable"', params$filtering_grouping_var, '".')
  }
}
cat(filtering_string)
```

## Exploratory analysis

```{r, echo=FALSE, results='asis'}
make_params_table('exploratory analysis', 'exploratory_', remove_pattern = TRUE)
```

## Differential analysis

```{r, echo=FALSE, results='asis'}
if (params$study_type == 'rnaseq'){
  make_params_table('DESeq2', 'deseq2_', remove_pattern = TRUE)
}
make_params_table('downstream differential analysis', 'differential_', remove_pattern = TRUE)
```

<!-- If any gene set methods have been activated show their params -->

```{r, echo=FALSE, results='asis'}
possible_gene_set_methods <- c('gsea')

if (any(unlist(params[paste0(possible_gene_set_methods, '_run')]))){
  cat("\n### Gene set analysis\n")

  for (gene_set_method in possible_gene_set_methods){
    if (unlist(params[paste0(gene_set_method, '_run')])){
      cat("\n### ", toupper(gene_set_method) ,"  {.tabset}\n")
      make_params_table(toupper(gene_set_method), paste0(gene_set_method, '_'), remove_pattern = TRUE)
    }
  }

}
```

# Appendices

## All parameters

```{r, echo=FALSE, results='asis'}
print( htmltools::tagList(datatable(params_table, caption = "All parameters", rownames = FALSE) ))
```

## Software versions

**Note:** For a more detailed accounting of the software and commands used (including containers), consult the execution report produced as part of the 'pipeline info' for this workflow.

```{r, echo=FALSE, results='asis'}
versions_table <- data.frame(do.call(rbind, strsplit(names(versions), split = '\\.')), unlist(versions))
colnames(versions_table) <- c('Component', 'Software', 'Version')
print( htmltools::tagList(datatable(versions_table, caption = "Software versions", rownames = FALSE, options = list(dom = 'ft', paging = FALSE)) ))
```

```{r, echo=FALSE, results='asis'}
htmltools::includeMarkdown(params$citations)
```
